\texttt{TREE}: A package for modelling plant TRait Ecology and Evolution

### 2015-05-28 ms.tex
Incorporate feedback from Sean
Clsoes #21

\mstype{Applications Note}
\runninghead{TREE}
% TODO: need more keywords here
\keywords{demography}

@@

\textbf{Key-words}: demography, emergent property, fitness, growth,
physiology, metapopulation, mortality, reproduction, size-structure,
tradeoff

@@

Plant growth and demography are fundamentally size- and trait- structured,
influencing dynamics over time-scales ranging from instantaneous physiological
effects to long-term evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As
an individual increases its leaf area, its potential to generate
photosynthate also rises. Yet as individuals get larger, an increasing
fraction of photosynthate must be diverted towards
building and maintaining support tissues rather than leaves
\citep{Givnish-1988, Enquist-2007}, and reproduction \citep{Thomas-2011}. Consequently 
rates of growth and mortality change with individual size
\citep{Muller-2006, Thomas-2011, Ruger-2011, Wright-2010}. The traits of
a species also interact with size to shape its demography. Leaf
construction costs and wood density affect growth rates
\citep{Wright-2010}. Other traits such as seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}. Over longer time-frames, natural selection favours
traits that provide a competitive advantage, such as faster growth, larger
seeds or taller height \citep{Falster-2003}. Consequently, selective forces may amplify size-related effects on demography. Accounting for the effects of size and traits in models is therefore fundamental to developing theory on the structure and dynamics of the biosphere
\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular for notable
phenomena such as successional dynamics, self thinning, trait evolution
and species coexistence.

Although the importance of size-structure for vegetation function and
demography has been long been recognised
\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001,
Kohyama-1993, Enquist-2007, Pacala-1996, Coomes-2007},
most large-scale mechanistic models of plant growth either omit it
entirely, or only include coarse approximations
\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
reason is the computational
challenge of modelling size-structured dynamics. During the 1980s and
early 1990s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
parameterised in terms of growth and demography, and consequently did a
poor job of simulating carbon fluxes \todo{ref}. The next generation of models
focussed more on physiology and, presumably for computational reasons,
sacrificed details of demography in order to achieve global coverage
\citep{Haxeltine-1996}. This history has brought us to the
current perplexing situation where all but a couple
\citep{Moorcroft-2001, Smith-2014} of the most popular vegetation
models lack size-structure. Instead, most models
are focused on modelling fluxes carbon and water, but say little about
demographic behaviour, and consequently are also unable to account for
any non-linear demographic feedbacks that may arise.

Interestingly, most models used in theoretical ecology to explore
questions about species coexistence also lack size-structured dynamics
\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013, Levin-1974,MacArthur-1967,Tilman-1985}.
It is common to assume competitive interactions are influenced by
size-related traits such as adult size and seed size, but detailed
size-structured demography is rarely considered, at least for plant
models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has
been difficult to connect mechanisms described in these simplistic
models with field data and real traits, which are almost always strongly
size-structured.

In this note, we describe the \texttt{TREE} package for R
\citep{R-2015}, a mechanistic framework for studying the effects of size
structure and trait variation on the demography of individual plants, of
patches of competing plants, and of meta-populations structured by a
prevailing disturbance regime. The package came about as a way to extend
previous work investigating the effect of traits on vegetation
properties \citep{Falster-2011} and the mechanisms facilitating
coexistence of trait mixtures \citep{Falster-2015}. These studies
provided the foundation methods needed \todo{needed for what.. its hinted above... maybe rearrange the sentences} here we provide robust and
extensible implementations. We describe the general approach of the
package and then a series of use cases, focussing at different levels of
ecological organisation. Each section provides a short description of
the required methods, design considerations, and then some worked
examples illustrating potential use cases and design features. Detailed
technical documents are provided as supplementary information, and updated
versions of technical documents are also available within the \texttt{TREE}
package itself. The use cases highlighted reflect our own interests in
understanding trait-based demography and community assembly. Yet we
expect the methods provided to be useful in other contexts. For example,
prior to \texttt{TREE} there was no \texttt{R} package that allowed one to simply grow a
plant over time, based on well-understood physiology. Surely this is one
of the simplest demographic actions one might hope for, with diverse
potential applications in ecology, evolution, forestry, agriculture, and
vegetation modelling. We hope the mechanistic platform provided here
will complement the increasingly sophisticated statistical methods now
available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.

@@

The default physiological model used in \texttt{TREE} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for diameter
growth also to be estimated. To properly model trait variation,
different parameters (such as rates of leaf turnover, respiration, photosynthesis 
per leaf area) in the model are linked via assumed tradeoffs. For
example, the trait LMA is used to estimate the rate of leaf turnover,
based on widely observed scaling relationship from \citet{Wright-2004}.

@@

\subsection{Use cases and examples}

@@

When modelling a patch, we are interested in modelling changes in the
size-density-distribution \(n(h,a)\) over time, i.e.~the density of
individuals with height \(h\) within a patch of age \(a\). We assume
that patches are large, and are vertically but not horizontally
structured. Thus, shading in \texttt{TREE} is modelled as a mean field with
respect to spatial layout. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Moorcroft-2001, Huston-1987, Smith-2014}. Under these
assumptions, the dynamics of \(n\) behave deterministically and can be
approximated via a Partial Differential Equation (PDE) (see Appendix
\ref{sec:demography}). The same PDE has been shown to capture average behaviour
across a large number of small patches \citep{Moorcroft-2001}. Ideally
one would also consider spatial interactions within patches,
however such models are very computationally demanding, and also
impossible to solve in a deterministic fashion \citep{Pacala-1996}.
Moreover, it remains unclear whether resolving spatial details within a
patch would provide further ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

@@

Beyond its conceptual appeal, this idea of a dynamic equilibrium allows
us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits, competitive
interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of patches. Unlike
some vegetation models, there is no need to set the density of seeds or
seedlings, these arise as outputs from the model. Characteristics of the
vegetation such as total leaf area, biomass, productivity and allocation
to different plant parts also arise as emergent properties (Fig.
\ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
growth rate with respect to size emerge at both patch scale and scale of
metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of relationships, dependent on patch
age. These emergent phenomena might then be compared to the observed
distributions in large forest tracts \citep{Muller-2006}.

@@

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available literature (see
Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
readily available. Providing updated parameter sets is trivial (see Appendix
\ref{sec:parameters}). Ideally, parameters would be inferred through a combination of
independent estimates and constraining outputs from \texttt{TREE} against diverse data sources
\citep{Lebauer-2012, Keenan-2013}.

>>>>

\mstype{Applications Note}
\runninghead{TREE}
\keywords{demography, emergent, fitness, growth,
physiology, metapopulation, mortality, reproduction, size-structure,
tradeoff}

@@

@@

Plant growth and demography are fundamentally size- and trait- structured,
influencing dynamics over time-scales ranging from instantaneous physiological
effects to long-term evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As
an individual increases its leaf area, its potential to generate
photosynthate also rises. Yet as individuals get larger, an increasing
fraction of that photosynthate is diverted towards
building and maintaining support tissues
\citep{Givnish-1988, Enquist-2007} and reproduction \citep{Thomas-2011}. Consequently 
rates of growth and mortality change with individual size
\citep{Muller-2006, Thomas-2011, Ruger-2011, Wright-2010}. The traits of
a species also interact with size to shape its demography. Leaf
construction costs and wood density affect growth rates
\citep{Wright-2010}. Other traits such as seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}. Over longer time-frames, natural selection generates
traits that amplify size-related effects on demography,
such as faster growth, larger seeds or taller height \citep{Falster-2003}. 
Accounting for the effects of size and traits in models is therefore fundamental to developing theory on the structure and dynamics of the biosphere
\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular for notable
phenomena such as successional dynamics, self thinning, trait evolution
and species coexistence.

Although the importance of size-structure for vegetation function and
demography has been long been recognised
\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001,
Kohyama-1993, Enquist-2007, Pacala-1996, Coomes-2007},
most large-scale mechanistic models of plant growth either omit it
entirely, or only include coarse approximations
\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
reason is the computational
challenge of modelling size-structured dynamics. During the 1980s and
early 1990s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
parameterised in terms of growth using statistical functions; this meant 
that data had to be collected for each species, limiting widespread application. 
The next generation of models focussed on establishing core physiology for 
a limited number of functional types, and, presumably for computational reasons,
sacrificed details of demography in order to achieve global coverage
\citep{Haxeltine-1996,Woodward-1995}. This history has brought us to the
current perplexing situation where all but a couple
\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular vegetation
models lack size-structure. Instead, most models
are focused on modelling fluxes carbon and water, but say little about
demographic behaviour, and consequently are also unable to account for
any non-linear demographic feedbacks that may arise.

Interestingly, most models used in theoretical ecology to explore
questions about species coexistence also lack size-structured dynamics
\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013, Levin-1974,MacArthur-1967,Tilman-1985}.
It is common to assume competitive interactions are influenced by
size-related traits such as adult size and seed size, but detailed
size-structured demography is rarely considered, at least for plant
models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has
been difficult to connect mechanisms described in these simplistic
models with type of data often collected in the field, such as physiological
leaf traits, and demographic data on growth and mortality in long-term monitoring 
plots \citep{Adler-2013}.

In this note, we describe the \texttt{TREE} package for R
\citep{R-2015}, a mechanistic framework for studying the effects of size
structure and trait variation on the demography of individual plants, of
patches of competing plants, and of meta-populations structured by a
prevailing disturbance regime. The package came about as a way to extend
previous work investigating the effect of traits on vegetation
properties and the mechanisms facilitating
coexistence of trait mixtures \citep{Falster-2011,Falster-2015}. Building 
on earlier studies \citep{Deroos-1997,Kohyama-1993,Moorcroft-2001}, these
papers outlined methods for driving trait-, size- and patch-structured 
dynamics from a core physiological model. Here we provide 
a robust and extensible implementation of those methods. 
We describe the general approach of the
package and then sections with details at different levels of
ecological organisation. Each section provides a short description of
the required methods, design considerations, and then 
example applications. Detailed
technical documents are provided as supplementary information (see Appendices),
and updated versions of technical documents are also available within the \texttt{TREE}
package itself.

While the applications highlighted reflect our own interests in understanding
trait-based demography and community assembly, we expect the methods provided
to be useful in variety of contexts. For example, prior to \texttt{TREE} there
was no \texttt{R} package that allowed one to simply grow a plant over time,
based on well-understood physiology. Surely this is one of the simplest
demographic actions one might hope for, with diverse potential applications in
ecology, evolution, forestry, agriculture, and vegetation modelling. More
specifically, \texttt{TREE} provides a  platform for investigating how
morphological and physiological traits influence  growth rate and shade
tolerance, thus potentially giving a theoretical foundation  for observed
empirical patterns \citep{Wright-2010,Baltzer-2007}. Being driven by traits,
the model can be extended to potentially very many species. Users can also 
add their own physiological models. Individual plants
can then be  situated in a patch with competitors and the demography of a
competing population  investigated, thereby capturing phenomena such as multi-
species self-thinning,  successional transitions and age-related changes in
vegetation productivity. \texttt{TREE}  models the emergent behaviour of
patches deterministically, based on a continual  size-density distribution. As
such, it is not required to specify a particular patch size.  Finally,
\texttt{TREE} allows for emergent phenomena across a metapopulation of patches
to  be investigated, and thus provides a tractable mechanism to scale from
individual plant to  vegetation, incorporating the non-linear effects of size
and traits on vegetation outcomes. In  this sense \texttt{TREE} is similar to
the Ecosystem Demography \citep{Moorcroft-2001}  and LPJ-GUESS
\citep{Smith-2014} models. However, whereas these other models were designed
to primarily capture large-scale regional biogeocehcmisty driven by climatic
variables,  \texttt{TREE} was designed so that the rules driving the
demography of individual trees, patches and metapopulations could be
manipulated and studied in detail. The three models are therefore highly 
complimentary, and also connect with the increasingly sophisticated statistical 
methods now available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.

@@

The default physiological model used in \texttt{TREE} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for diameter
growth also to be estimated. To properly model trait variation,
different parameters (such as rates of leaf turnover, respiration, photosynthesis 
per leaf area) in the model are linked via assumed tradeoffs. For
example, the trait LMA is used to estimate the rate of leaf turnover,
based on observed scaling relationship spanning across diverse vegetation and
plant functional  global \citep{Wright-2004} (see 
Appendix \ref{sec:FFW16} for details).

@@

\subsection{Applications}

@@

When modelling a patch, we are interested in modelling changes in the
size-density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(h\) within a patch of age \(a\). Density
is measured as individuals per unit height and per unit ground area.

We assume that patches are large, and are vertically but not horizontally
structured. In other words, we account for size differences but not spatial 
layout within the patch. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Moorcroft-2001, Huston-1987, Smith-2014,Kohyama-1993}. The assumption of a large
patch means the population of plants within the patch -- and thus the 
dynamics of \(n\) -- behave deterministically \citep{Deroos-1997}, and can therefore be
approximated via a Partial Differential Equation (PDE) (see Appendix
\ref{sec:demography}). The same PDE has been shown to capture average behaviour
across a large number of small patches \citep{Moorcroft-2001}. The model may 
therefore capture average demographic behaviour across a range of patch sizes.

Ideally one would also consider spatial interactions within patches,
however such models are very computationally demanding, and also
impossible to solve in a deterministic fashion \citep{Huston-1987, 
Pacala-1996}.
Moreover, it remains unclear whether resolving spatial details within a
patch would provide further ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

@@

Beyond its conceptual appeal, this idea of a dynamic equilibrium allows
us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits, competitive
interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of patches. Unlike
some vegetation models, there is no need to set the density of seeds or
seedlings, these arise as outputs from the model. Characteristics of the
vegetation such as total leaf area, biomass, productivity and allocation
to different plant parts also arise as emergent properties (Fig.
\ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
growth rate with respect to size emerge at both patch scale and scale of
metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of height-density relationships, 
dependent on patch age, as well as an emergent average across all patches 
(Fig. \ref{fig:emergent}). This emergent relationship might then be 
compared to the observed distributions in large forest plots \citep{Muller-2006}.

@@

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available literature (see
Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
readily available. Providing updated parameter sets is trivial; the challenge 
is to infer suitable values by constraining  outputs from \texttt{TREE} against diverse data sources
\citep{Lebauer-2012, Keenan-2013}. This is an area in which we are still actively working.
